---
title: "Pizza Sales Analysis & Revenue Forecasting"
author: "Jonathan Katz"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5
)
```

# Introduction

**Business Context:** A pizza restaurant wants to better understand their sales patterns and forecast future revenue to improve staffing, inventory, and operations.

**Dataset:** 48,620 pizza orders from 2015 (full year)

**Questions We're Answering:**

1. What will total sales revenue be for the next 7 days?
2. What are the peak hours for orders?
3. Which day of the week has the highest sales?
4. What is the best-selling pizza?
5. What is the most popular pizza size?
6. What is the average order value?

---

# Setup & Load Data

```{r load-libraries}
# Load libraries
library(tidyverse)
library(lubridate)
library(scales)
library(readxl)
library(knitr)
library(kableExtra)

# Set theme for all plots
theme_set(theme_minimal(base_size = 12))
```

```{r load-data}
# Load data
pizza <- read_excel("../data/raw/Data Model - Pizza Sales.xlsx")

# Quick look at the data
glimpse(pizza)
```

```{r clean-data}
# Clean and prepare data
pizza <- pizza %>%
  mutate(
    order_date = as.Date(order_date),
    order_hour = hour(hms(order_time)),
    day_of_week = wday(order_date, label = TRUE, week_start = 1),
    month = month(order_date, label = TRUE)
  )

# Confirm data range
cat("Date range:", as.character(min(pizza$order_date)), "to", as.character(max(pizza$order_date)), "\n")
cat("Total orders:", format(n_distinct(pizza$order_id), big.mark = ","), "\n")
cat("Total revenue: $", format(sum(pizza$total_price), big.mark = ","), "\n")
```

---

# Question 1: What will total sales revenue be for the next 7 days?

*This question will be answered in the forecasting notebook (02_forecasting.Rmd). Here we prepare the daily sales data.*

```{r daily-sales-prep}
# Aggregate daily sales for forecasting
daily_sales <- pizza %>%
  group_by(order_date) %>%
  summarise(
    total_revenue = sum(total_price),
    total_orders = n_distinct(order_id),
    total_pizzas = sum(quantity),
    .groups = "drop"
  )

# Preview daily sales
daily_sales %>%
  head(10) %>%
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r daily-sales-trend}
# Plot daily revenue trend
daily_sales %>%
  ggplot(aes(x = order_date, y = total_revenue)) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  geom_smooth(method = "loess", color = "red", se = FALSE, linewidth = 1) +
  labs(
    title = "Daily Revenue Trend (2015)",
    subtitle = "Blue = actual, Red = trend line",
    x = "Date",
    y = "Daily Revenue ($)"
  ) +
  scale_y_continuous(labels = dollar) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")
```

**Observation:** Revenue is relatively stable throughout the year with some weekly variation. No strong upward or downward trend.

---

# Question 2: What are the peak hours for orders?

```{r peak-hours}
# Orders by hour
hourly_orders <- pizza %>%
  group_by(order_hour) %>%
  summarise(
    total_orders = n_distinct(order_id),
    total_revenue = sum(total_price),
    .groups = "drop"
  )

# Plot
hourly_orders %>%
  ggplot(aes(x = order_hour, y = total_orders)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = comma(total_orders)), vjust = -0.5, size = 3) +
  labs(
    title = "What are the peak hours for orders?",
    subtitle = "Total orders by hour of day (2015)",
    x = "Hour of Day",
    y = "Number of Orders"
  ) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1)))
```

```{r peak-hours-summary}
# Top 3 peak hours
top_hours <- hourly_orders %>%
  arrange(desc(total_orders)) %>%
  head(3)

cat("Peak Hours:\n")
cat("1st:", top_hours$order_hour[1], ":00 -", comma(top_hours$total_orders[1]), "orders\n")
cat("2nd:", top_hours$order_hour[2], ":00 -", comma(top_hours$total_orders[2]), "orders\n")
cat("3rd:", top_hours$order_hour[3], ":00 -", comma(top_hours$total_orders[3]), "orders\n")
```

**Answer:** Peak hours are **12pm (lunch)** and **5-7pm (dinner)**. The restaurant should ensure full staffing during these windows.

---

# Question 3: Which day of the week has the highest sales?

```{r day-of-week}
# Sales by day of week
dow_sales <- pizza %>%
  group_by(day_of_week) %>%
  summarise(
    total_revenue = sum(total_price),
    total_orders = n_distinct(order_id),
    avg_daily_revenue = total_revenue / 52,  # 52 weeks in a year
    .groups = "drop"
  )

# Plot
dow_sales %>%
  ggplot(aes(x = day_of_week, y = total_revenue, fill = total_revenue)) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  geom_text(aes(label = dollar(total_revenue, accuracy = 1)), vjust = -0.5, size = 3.5) +
  scale_fill_gradient(low = "lightblue", high = "steelblue") +
  labs(
    title = "Which day of the week has the highest sales?",
    subtitle = "Total revenue by day of week (2015)",
    x = "Day of Week",
    y = "Total Revenue ($)"
  ) +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.1)))
```

```{r dow-summary}
# Best and worst days
best_day <- dow_sales %>% filter(total_revenue == max(total_revenue))
worst_day <- dow_sales %>% filter(total_revenue == min(total_revenue))

cat("Best day:", as.character(best_day$day_of_week), "- $", comma(best_day$total_revenue), "\n")
cat("Worst day:", as.character(worst_day$day_of_week), "- $", comma(worst_day$total_revenue), "\n")
cat("Difference:", dollar((best_day$total_revenue - worst_day$total_revenue) / worst_day$total_revenue * 100, accuracy = 0.1, prefix = ""), "% more revenue on", as.character(best_day$day_of_week))
```

**Answer:** **Friday** has the highest sales, followed by Thursday and Saturday. Sunday has the lowest sales — consider reduced hours or staffing.

---

# Question 4: What is the best-selling pizza?

```{r best-selling-pizza}
# Top 10 pizzas by quantity sold
top_pizzas <- pizza %>%
  group_by(pizza_name) %>%
  summarise(
    total_sold = sum(quantity),
    total_revenue = sum(total_price),
    .groups = "drop"
  ) %>%
  arrange(desc(total_sold)) %>%
  head(10)

# Plot
top_pizzas %>%
  ggplot(aes(x = reorder(pizza_name, total_sold), y = total_sold)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = comma(total_sold)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(
    title = "What is the best-selling pizza?",
    subtitle = "Top 10 pizzas by quantity sold (2015)",
    x = "",
    y = "Total Pizzas Sold"
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15)))
```

```{r best-pizza-summary}
cat("Best-selling pizza:", top_pizzas$pizza_name[1], "\n")
cat("Total sold:", comma(top_pizzas$total_sold[1]), "pizzas\n")
cat("Total revenue: $", comma(top_pizzas$total_revenue[1]), "\n")
```

**Answer:** **The Classic Deluxe Pizza** is the best-seller with over 2,400 pizzas sold, generating significant revenue.

---

# Question 5: What is the most popular pizza size?

```{r pizza-size}
# Sales by size
size_sales <- pizza %>%
  group_by(pizza_size) %>%
  summarise(
    total_sold = sum(quantity),
    total_revenue = sum(total_price),
    pct_of_sales = total_sold / sum(pizza$quantity) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(total_sold))

# Reorder size factor
size_order <- c("S", "M", "L", "XL", "XXL")
size_sales <- size_sales %>%
  mutate(pizza_size = factor(pizza_size, levels = size_order))

# Plot
size_sales %>%
  ggplot(aes(x = pizza_size, y = total_sold, fill = pizza_size)) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  geom_text(aes(label = paste0(comma(total_sold), "\n(", round(pct_of_sales, 1), "%)")),
            vjust = -0.3, size = 3.5) +
  labs(
    title = "What is the most popular pizza size?",
    subtitle = "Total pizzas sold by size (2015)",
    x = "Pizza Size",
    y = "Total Pizzas Sold"
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Blues")
```

```{r size-summary}
top_size <- size_sales %>% filter(total_sold == max(total_sold))
cat("Most popular size:", top_size$pizza_size, "\n")
cat("Percentage of all sales:", round(top_size$pct_of_sales, 1), "%\n")
```

**Answer:** **Large (L)** is the most popular size, accounting for nearly half of all pizzas sold. Consider optimizing inventory and pricing around this size.

---

# Question 6: What is the average order value?

```{r avg-order-value}
# Calculate order-level metrics
order_summary <- pizza %>%
  group_by(order_id, order_date) %>%
  summarise(
    order_total = sum(total_price),
    pizzas_per_order = sum(quantity),
    .groups = "drop"
  )

# Summary statistics
avg_order_value <- mean(order_summary$order_total)
median_order_value <- median(order_summary$order_total)
avg_pizzas <- mean(order_summary$pizzas_per_order)

cat("Average Order Value: $", round(avg_order_value, 2), "\n")
cat("Median Order Value: $", round(median_order_value, 2), "\n")
cat("Average Pizzas per Order:", round(avg_pizzas, 2), "\n")
```

```{r aov-distribution}
# Distribution of order values
order_summary %>%
  ggplot(aes(x = order_total)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = avg_order_value, color = "red", linewidth = 1, linetype = "dashed") +
  annotate("text", x = avg_order_value + 10, y = 2000,
           label = paste("Avg: $", round(avg_order_value, 2)), color = "red") +
  labs(
    title = "What is the average order value?",
    subtitle = "Distribution of order totals (2015)",
    x = "Order Total ($)",
    y = "Number of Orders"
  ) +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = comma)
```

**Answer:** The average order value is **$38.31**. Most orders fall between $15-$50. This metric is useful for setting promotions (e.g., "Free delivery on orders over $40").

---

# Summary of Findings

```{r summary-table}
# Create summary table
findings <- tibble(
  Question = c(
    "Peak hours?",
    "Best day of week?",
    "Best-selling pizza?",
    "Most popular size?",
    "Average order value?"
  ),
  Answer = c(
    "12pm (lunch), 5-7pm (dinner)",
    "Friday",
    "The Classic Deluxe Pizza",
    "Large (L) - 46% of sales",
    "$38.31"
  ),
  Business_Action = c(
    "Full staffing during peak hours",
    "Run promotions on slow days (Sunday)",
    "Ensure Classic Deluxe ingredients always stocked",
    "Optimize Large pizza pricing and inventory",
    "Set delivery minimum at $35-40"
  )
)

findings %>%
  kable(col.names = c("Question", "Answer", "Recommended Action")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"))
```

---

# Next Steps

The next notebook (**02_forecasting.Rmd**) will answer:

> **"What will total sales revenue be for the next 7 days?"**

We'll build a time series forecast using the `daily_sales` data prepared above.

---

*Analysis completed on `r Sys.Date()`*
